---
title: "Homework 1"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "2025-02-06"
---
```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
## Part 1
### 1.1
load libraries
```{r}
library(dplyr)  
library(lubridate) 
```
- load dataset
- print first 5 rows
```{r}
crime_data <- read.csv("crime_data.csv", stringsAsFactors = FALSE)

head(crime_data, 5)
```
### 1.2
- get number of missing values for columns
- delete columns which miss more than 50% of data
```{r}
missing_values <- colSums(is.na(crime_data))

missing_values[missing_values > 0]

threshold <- 0.5 * nrow(crime_data)
columns_to_drop <- names(missing_values[missing_values > threshold])

crime_data_cleaned <- crime_data %>% select(-one_of(columns_to_drop))

names(crime_data_cleaned)

```
### 1.3
- Convert Date.OCC to date format
- Extract Year, Month, Day to new columns
- Calculate Hour from TIME.OCC
```{r}
crime_data_cleaned$DATE.OCC <- mdy_hms(crime_data_cleaned$DATE.OCC)

crime_data_cleaned$Year <- year(crime_data_cleaned$DATE.OCC)
crime_data_cleaned$Month <- month(crime_data_cleaned$DATE.OCC)
crime_data_cleaned$Day <- day(crime_data_cleaned$DATE.OCC)

crime_data_cleaned$Hour <- as.integer(crime_data_cleaned$TIME.OCC / 100)

head(crime_data_cleaned[, c("DATE.OCC", "TIME.OCC", "Year", "Month", "Day", "Hour")])

```
### 1.4
- Filter for 2023
- Filter for burglaries
- check if size changed
```{r}
crime_data_2023 <- crime_data_cleaned %>% filter(Year == 2023)

crime_burglary_2023 <- crime_data_2023 %>% 
  filter(grepl("BURGLARY", crime_data_2023$Crm.Cd.Desc, ignore.case = TRUE))

cat("unfiltered data:", dim(crime_data_cleaned))
cat("data for 2023:", dim(crime_data_2023))
cat("data for burglary + 2023:", dim(crime_burglary_2023))
```
### 1.5
- Group by AREA.NAME
- Calculate total crimes and avg victim age
- Display results
```{r}
crime_summary <- crime_burglary_2023 %>%
  group_by(AREA.NAME) %>%
  summarise(
    Total_Crimes = n(),
    Avg_Victim_Age = mean(Vict.Age, na.rm = TRUE)
  ) %>%
  arrange(desc(Total_Crimes))

print(crime_summary, n = Inf)
```
## Part 3
### 3.1
- Group by Month
- summaries total crimes for each month
- Display results
```{r}
crimes_by_month <- crime_data_cleaned %>%
  group_by(Month) %>%
  summarise(Total_Crimes = n())

print(crimes_by_month)
```
### 3.2
- Filter crimes where weapon was not used
- Get number of such crimes
- Display results

<em> NOTE: Using original crime_data because crime_data_cleaned does not have the Weapon.Used.Cd column </em>
```{r}
crimes_with_weapon <- crime_data %>%
  filter(!is.na(Weapon.Used.Cd)) %>%  
  summarise(Weapon_Crimes = n())      

print(crimes_with_weapon)
```
### 3.3
- Group by premis.desc
- Get number of crimes for each premis.desc
- Display results
```{r}
crime_by_premis_desc <- crime_data_cleaned %>%
  group_by(Premis.Desc) %>%          
  summarise(Total_Crimes = n())       

print(crime_by_premis_desc)
```
## Part 4
- Add a severity.score column which will be based on the rows data
- Group by area and get sum of severity scores for each area
- Display results

<em> NOTE: Using original crime_data because crime_data_cleaned does not have the Weapon.Used.Cd column </em>
```{r}
crime_data <- crime_data %>%
  mutate(
    Severity.Score = case_when(
      !is.na(Weapon.Used.Cd) ~ 5,  
      grepl("BURGLARY", Crm.Cd.Desc, ignore.case = TRUE) ~ 3, 
      TRUE ~ 1  
    )
  )

severity_by_area <- crime_data %>%
  group_by(AREA.NAME) %>%
  summarise(Total_Severity_Score = sum(Severity.Score))

print(severity_by_area)
```